How to \emph{not} write Virtual Machines for Dynamic Languages

### 2007-07-15 talk/dyla2007/dyla.tex
My final review, 2nd half.

In the sequel, we will focus on the relative advantages and
inconveniences of the PyPy approach compared to the approach of
hand-writing a language implementation on top of an OO VM.

@@

Of course, the translation toolchain, once written, can also be reused
to implement other languages, and possibly tailored on a case-by-case
basis to fit the specific needs of a language.  The process is
incremental: we can add more features as needed instead of starting from
a maximal up-front design, and gradually improve the quality of the
tools, the garbage collectors, the various optimizations, etc.

@@

As mentioned above, the performance of the VMs generated by our
translation framework are quite acceptable -- e.g. the Python VM
generated via C code is much faster than Jython running on the best
JVMs.  Of course, the JIT compilers in these JVMs are essential to
achieve even this performance, which further proves the point that
writing good OO VMs -- especially ones meant to support dynamic
languages -- is a lot of work.

The deeper problem with the otherwise highly-tuned JIT compilers of the
OO VMs is that they are not a very good match for running dynamic
languages.  It might be possible to tune a general-purpose JIT compiler
enough and write the dynamic language implementation accordingly so
that most of the bookkeeping work involved in running the dynamic
language can be removed -- dispatching, boxing, unboxing...  However
this has not been demonstrated yet.\footnote
{Still in the draft stage, a proposed
extension to the Java bytecode \cite{invokedynamic} might help achieve
better integration between the Java JITs and dynamic language
implementations running on top of JVMs.}

By far the fastest Python implementation, Psyco \cite{psyco-software}
contains a hand-written language-specific dynamic compiler.  It works by
specializing (parts of) Python functions by feeding runtime information
back into the compiler (typically, but not exclusively, object types).
The reader is referred to \cite{Psyco-paper} for more details.

PyPy abstracts on this approach: its translation tool-chain is able to
extend the generated VMs with an \emph{automatically generated} dynamic
compiler that uses techniques similar to those of Psyco, derived from
the interpreter.  This is achieved by a pragmatic application of partial
evaluation techniques guided by a few hints added to the source of the
interpreter.  In other words, it is possible to produce a reasonably
good language-specific JIT compiler and insert it into a VM, alongside
with the necessary support code and the rest of the regular interpreter.

This result was one of the major goals and motivations for the whole
approach.  By construction, the JIT stays synchronized with its VM
and with the language when it evolves,
and any code written in the dynamic language runs
correctly under the JIT.  Some very simple Python examples run more than
100 times faster.  At the time of this writing this is still rather
experimental, and the techniques involved are well beyond the scope of
the present paper.  The reader is referred to \cite{D08.2} for more
information.

@@

%XXX doesn't look entirely nice
\begin{itemize}
\item \emph{High-level languages are suitable to implement dynamic languages.}
They allow an interpreter to be written more abstractly, which has many
advantages -- among them the avoidance of a proliferation of diverging
implementations, and better ways to combine flexibility with efficiency.
Moreover, this is not incompatible with targeting and benefiting from
existing high-quality object-oriented virtual machines like those of the
Java and .NET.

\item \emph{Do not write VMs ``by hand''.}
In other words, write an \emph{interpreter} but not a
\emph{virtual machine} for the language.
Writing language-specific virtual machines is a time-consuming task for
medium to large languages.  Unless large amounts of resources can be
invested, the resulting VMs are bound to have limitations which lead to
the emergence of many implementations, a fact that is taxing precisely
for a community with limited resources.  (This is of course even more
true for general-purpose VMs.)
\end{itemize}

@@

\begin{itemize}
\item \emph{Let's write more meta-programming translation toolchains.}
Aside from the advantages described in section
\ref{sect:metaprogramming}, a translation toolchain need not be
standardized for inter-operability but can be tailored to the needs of
each project.  Diversity is good; translation toolchains offset the need
to attempt to standardize on a single OO VM.
\end{itemize}

>>>>

In the sequel, we will focus on the relative advantages and
inconveniences of the PyPy approach compared to
hand-writing a language implementation on top of an OO VM.

@@

Of course, the translation toolchain, once written, can also be reused
to implement other languages, and tailored on a case-by-case
basis to fit the specific needs of a language.  The process is
incremental: we can add more features as needed instead of starting from
a maximal up-front design, and gradually improve the quality of the
tools, the garbage collectors, the various optimizations, etc.

@@

As mentioned above, the performance of the VMs generated by our
translation framework is quite acceptable -- e.g. the Python VM
generated via C code is much faster than Jython running on the best
JVMs.  Moreover, the JIT compilers in these JVMs are essential to
achieve even this performance, which further proves the point that
writing good general-purpose OO VMs -- especially ones meant to support dynamic
languages -- is a lot of work.

The deeper problem with the otherwise highly-tuned JIT compilers of the
OO VMs is that they are not a very good match for running arbitrary dynamic
languages.  It might be possible to tune a general-purpose JIT compiler
enough and write the dynamic language implementation accordingly so
that most of the bookkeeping work involved in running the dynamic
language can be removed -- dispatching, boxing, unboxing...  However
this has not been demonstrated yet.\footnote
{Still in the draft stage, a proposed
extension to the Java bytecode \cite{invokedynamic} might help achieve
better integration between the Java JITs and some class of dynamic languages
running on top of JVMs.}

By far the fastest Python implementation, Psyco \cite{psyco-software}
contains a hand-written language-specific dynamic compiler.  It works by
specializing (parts of) Python functions based on runtime information
fed back into the compiler (typically, but not exclusively, object types).
The reader is referred to \cite{Psyco-paper} for more details.

PyPy abstracts on this approach: its translation tool-chain is able to
extend the generated VMs with an \emph{automatically generated} dynamic
compiler that uses techniques similar to those of Psyco, derived from
the interpreter.  This is achieved by a pragmatic application of partial
evaluation techniques guided by a few hints added to the source of the
interpreter.  In other words, it is possible to produce a reasonably
good language-specific JIT compiler and insert it into a VM, along
with the necessary support code and the rest of the regular interpreter.

This result was one of the major goals and motivations for the whole
approach.  By construction, the JIT stays synchronized with its VM
and with the language when it evolves.  Also by construction, the JIT
immediately supports (and is correct for) arbitrary input code.
Some very simple Python examples run more than
100 times faster.  At the time of this writing this is still rather
experimental, and the techniques involved are well beyond the scope of
the present paper.  The reader is referred to \cite{D08.2} for more
information.

@@

%XXX doesn't look entirely nice
\begin{itemize}
\item \emph{High-level languages are suitable to implement dynamic languages.}
They allow an interpreter to be written more abstractly, which has many
advantages.  Among these, it avoids the proliferation of diverging
implementations, and gives implementers better ways to combine flexibility
with efficiency.
Moreover, this is not incompatible with targeting and benefiting from
existing high-quality object-oriented virtual machines like those of the
Java and .NET.

\item \emph{Do not write VMs ``by hand''.}
In other words, write an \emph{interpreter} but not a
\emph{virtual machine} for the language.
Writing language-specific virtual machines is a time-consuming task for
medium to large languages.  Unless large amounts of resources can be
invested, the resulting VMs are bound to have limitations which lead to
the emergence of many implementations, a fact that is taxing precisely
for a community with limited resources.  This is of course even more
true for VMs that are meant to be general-purpose.
\end{itemize}

@@

\begin{itemize}
\item \emph{Let's write more meta-programming translation toolchains.}
Aside from the advantages described in section
\ref{sect:metaprogramming}, a translation toolchain need not be
standardized for inter-operability but can be tailored to the needs of
each project.
\item \emph{Diversity is good.}  Meta-programming translation
toolchains offset the need for standardization of general-purpose OO VMs.
\end{itemize}
